---
title: "RepData_PeerAssignment1"
author: "Nick Hitt"
date: "11/14/2020"
output: html_document
keep_md: true
---


## Loading and preprocessing the data
```{r dataprocessing, include=FALSE}
fileName <- "Activity.zip"
dir <- "activity"

library(lubridate)
library(dplyr)
```
```{r dataprocessing2, include=TRUE}
if(!file.exists(dir)){
  unzip("Activity.zip", files = NULL, exdir=".")
}

activity <- read.csv("activity.csv") 

```

We need to remove the zeros from steps because the zeros bias the median calculation
```{r dataprocessing3, include=TRUE}
steps_zer_removed <- activity[which(activity$steps != 0),] # removing zeros from steps because the zeros bias the median calculation

```
## What is mean total number of steps taken per day?

First we need to calculate the average, sum and median steps and plot a histogram
```{r meansteps, include=TRUE}

daily_avg_steps <- aggregate(steps ~ date, activity, mean, na.rm = TRUE) 

med_steps <- aggregate(as.numeric(steps) ~ date, steps_zer_removed, median, na.rm = TRUE) 

sum_steps <- aggregate(steps ~ date, activity, sum, na.rm = TRUE) 

hist(sum_steps[,2], main = "Histogram of Daily Sum of Steps",
     xlab = "Sum of Steps", ylab = "Daily Frequency", xaxt = "n")
axis(side=1, at=(seq(0,25000,5000)), labels=seq(0,25000,5000))

sum_steps
```


## What is the average daily activity pattern?
Here we need to find the average amount of steps per time interval of the day
```{r avgactpattern, include=TRUE}

int_avg_steps <- aggregate(steps ~ interval, activity, mean, na.rm = TRUE)
```
Then we need to find what time of the day has the highest steps
```{r avgactpattern2, include=TRUE}

max_steps <- max(int_avg_steps[,2]) 
int_max_steps <- int_avg_steps[which(int_avg_steps[,2] == max_steps),1] 
```
Then plot the data
```{r avgactpattern3, include=TRUE}

library(ggplot2)
g <- ggplot(int_avg_steps,aes(interval,steps))

p<- g + geom_point() +geom_line() + ggtitle(expression("Total Steps by Interval")) +
  ylab(expression("Total Steps")) +
  xlab("Interval")

print(p)
```

## Imputing missing values
Now we need to 'fix' the data by replacing the missing values with data.
We will do this by running a for loop that checks if each data point is missing or not.
If it is missing then we will replace it with the average steps in that interval
```{r fixmiss, include=TRUE}

activity_na <- is.na(activity$steps) 
total_na <- sum(activity_na) 


fillNA <- numeric() 
for (i in 1:nrow(activity)) { 
  obs <- activity[i, ] 
  if (is.na(obs$steps)) { 
    steps <- subset(int_avg_steps, interval == obs$interval)$steps 
  } else { 
    steps <- obs$steps
  }
  fillNA <- c(fillNA, steps) 
}
```
Now we should check and see how that effected the data. We will do that
by making a histogram of thew daily sum of steps
```{r fixmiss2, include=TRUE}
new_activity <- activity 
new_activity$steps <- fillNA 

new_steps_zer_removed <- new_activity[which(new_activity$steps != 0),] 

new_daily_avg_steps <- aggregate(steps ~ date, new_activity, mean, na.rm = TRUE) 
new_sum_steps <- aggregate(steps ~ date, new_activity, sum, na.rm = TRUE) 
new_med_steps <- aggregate(steps ~ date, new_activity, median, na.rm = TRUE) 


hist(new_sum_steps[,2], main = "Histogram of Daily Sum of Steps",
     xlab = "Sum of Steps", ylab = "Daily Frequency", xaxt = "n")
axis(side=1, at=(seq(0,25000,5000)), labels=seq(0,25000,5000))
```

## Are there differences in activity patterns between weekdays and weekends?
Now we should check and see if there are differences between activity patterns
between the weekend and week days. We will change the dates to weekdays/weekends
and then use those to plot a panel plot of the data to compare
```{r weekvsweekend, include=TRUE}

new_activity$date <- as.Date(new_activity$date) 
days <- c(1:nrow(new_activity)) 

factor_days <- c(1:nrow(new_activity)) 
days <- weekdays(new_activity$date)

days <- factor(days, levels = c("Monday", "Tuesday" , "Wednesday", "Thursday", "Friday","Saturday", "Sunday"),
             labels = c("Weekday","Weekday","Weekday","Weekday","Weekday","Weekend","Weekend"))

days <- data.frame("days" = days) 

new_activity <- cbind(new_activity, days) 

day_steps <- aggregate(steps ~ interval + days, new_activity, mean, na.rm = TRUE) 

g1 <- ggplot(data = new_activity, aes(steps, interval)) +
  geom_line(color = "steelblue", size = 1) +
  labs(y = "Number of Steps", x = "Interval") + 
  facet_grid(days ~.)

print(g1)
```

